Narrative

### 2023-05-25 manuscript/manuscript.qmd
manuscript: added discussion to qmd

>>>>

# Discussion

We used Monte-Carlo simulations to study the total bias that should be expected in meta-analytic heterogeneity estimates due to unreliability in primary studies. Our results indicate that uncontrolled unreliability in primary studies should under almost all circumstances be expected to lead to an severe underestimate of heterogeneity in meta-analysis. We observed a positive bias in heterogeneity estimates due to unreliability only under zero heterogeneity, or under exceptional circumstances (i.e., unusual combinations of low heterogeneity, large variance in primary study reliabilities, large effect sizes, and low number of meta-analyzed studies). This generally negative bias was observed under conditions that should have been favorable to a positive bias (i.e., large variance in primary study reliability), indicating that negative bias should probably be expected in all cases except for zero true heterogeneity. Our results were approximately the same whether one used a classic random-effects meta-analytic model (Hedges & Vevea, 1998) or a 'bare-bones' psychometric meta-analytic model (Hunter & Schmidt, 2015), and whether one used Pearson's $r$ as the effect size or Fisher's $z$, although the negative bias was slightly larger with Fisher's $z$. The severity of the underestimation of heterogeneity suggests meta-analyses in psychology often miss moderators and a risk that theoretical development in psychology has been misguided by meta-analyses.

The negative bias in heterogeneity estimates can be extreme if reliability is low. For a typical effect size around $r$ = 0.2, larger heterogeneity ($\tau > 0.1$) can be expected to be underestimated by about 15% - 60% primarily depending on average reliability in primary studies (higher reliability leads to a smaller underestimate). The underestimate is worse with lower average effect size, but although it improves with larger effect sizes it is likely to remain negative, and less variability in primary study reliabilities will lead to larger negative bias. As true heterogeneity increases, the proportional underestimate decreases to some degree, but the underestimate in absolute terms will worsen. Although small degrees of true heterogeneity ($\tau < .08$) can be overestimated under certain conditions, this would typically require a combination of a small number of studies, large variance in study reliabilities, and a large average effect size. The severity and generality of the underestimation means that meta-analysts cannot ignore the effect of reliability on their heterogeneity estimates.

If true heterogeneity is zero, there is a positive bias in observed heterogeneity estimates, but for even small degrees of true heterogeneity there tends to be a negative bias in observed estimates. As such, it is not possible to know whether a small degree of observed heterogeneity is due to overestimated true zero heterogeneity or underestimated small heterogeneity. However, for observed heterogeneity above $\tau > 0.025$ one can usually be confident that the estimate is an underestimate of true heterogeneity rather than an overestimate of zero heterogeneity if the average effect sizes is below approximately $r = 0.2$.

Similarly, if researchers find zero heterogeneity in their observed estimates, true heterogeneity is most likely to be zero or small. Larger true heterogeneity (e.g., $\tau > 0.08$) has a higher expected observed heterogeneity value even when average reliability is as low as 0.6, and as such is expected to result in zero estimates less often than zero or smaller true heterogeneity for fixed conditions. For example, for the conditions reported in the main article (i.e., K = 20, N = 150, $\sigma_R$ = 0.15), an assuming a typical (reported) average reliability of 0.8 and a zero average correlation, then an observed zero heterogeneity estimate is XX times more likely to correspond to true zero heterogeneity than $\tau = 0.1$. That said, an observed zero heterogeneity estimate is still possible to observe under larger true heterogeneity levels as variability in heterogeneity estimates is often large in practice (Ioannidis; ref2). In addition, separating between zero and small heterogeneity is unfeasible considering realistic sample sizes in meta-analyses in psychology (Olsson-Collentine et al., 2020). In our view, the distinction between zero and small heterogeneity in meta-analysis is primarily a hypothetical one, as we consider low heterogeneity and meta-analysis of a large number of studies unlikely to cooccur in practice.

We found that using Pearson's $r$ or Fisher's $z$ as our effect size resulted in very similar degrees of bias in heterogeneity estimates due to uncorrected unreliability in primary studies. The underestimate is slightly worse for Fisher's $z$ than for Pearson's $r$. The usage of Pearson's $r$ adds a slight positive bias to heterogeneity estimates (due to its bounded nature and effect size dependent sampling variance) which apparently is beneficial for heterogeneity estimates when not correcting for unreliability. However, the additional negative bias when using Fisher's $z$ is small and we consider the difference negligible in the context of the severe underestimates generated by unreliability.

## Should meta-analysts correct for unreliability in primary studies?

Despite the severity of underestimation due to unreliability, it is neither straightforward to correct for unreliability using the formulas of classical test theory nor self-evident that doing so is the optimal solution. Although our simulation results are not affected by whether researchers are interested in observed measures or latent constructs, it is safe to assume that the majority of researcher in psychology are interested in latent constructs. A central assumption behind applying the classical test theory correction to examine latent constructs is that disattenuated effect sizes (“true scores”) correspond approximately to latent construct scores (Hunter & Schmidt, 2015; Borsboom & Mellenbergh, 2002). Given that these assumptions hold, or one is satisfied with correcting for measurement error regardless of the possibility of a latent construct interpretation, there are several complexities that arise when attempting to correct for unreliability in practice. First, reliability estimates are lower bound estimates, meaning systematically correcting for unreliability can be expected to lead to overestimated effect sizes and heterogeneity. Unfortunately, the most reported reliability estimate (Chronbach’s alpha; Flake et al., 2017) is known to also underestimate the lower bound (Sijtsma, YEAR). Second, estimates of reliability (e.g., those based on a single measurement occasion such as Cronbach’s alpha) do not necessarily account for all types of measurement error (Hunter & Schmidt, 2015, p. XX, section “Using the appropriate reliability coefficient”). If so, they overestimate reliability. Hence, from a meta-analytic perspective, most reliability estimates available for correcting effect sizes are likely to be affected by both a positive and a negative bias compared to the true reliability, but what the total effect is likely to be is uncertain (and hence how accurate a heterogeneity estimate based on these corrections would be). Even so, in view of these two problems one could argue that an imperfect correction of unreliability is likely to be preferable over no correction (e.g., Oswald, YEAR), especially given the severe underestimation of heterogeneity. Further study would be useful to provide evidence to whether this is truly the case.

Another concrete problem is that reliability estimates are sometimes not reported in primary studies. Missing reliability estimates must then be imputed, which is usually done by taking the average of observed reliabilities (e.g., Gnambs & Sengewald. 2023) or imputing some ‘typical’ reliability (Hunter & Schmidt, 2015, p. XX). Both approaches have problems. There can be much debate about what a ‘typical’ reliability is (LeBreton et al., 2014) and selecting different values can have large effect on estimates. Mean imputation is viewed as inappropriate by experts in missing data (e.g., Enders, 2010, p. 42-43; van Buuren, 2018, p. 12; Schafer, 2002, p. 159) partly because the pooled estimate will be biased unless observed values coincide with the population mean, particularly so if missing values are related to the value of the statistic (and it seems highly likely that missingness is more common with low reliability). However, it is not clear that more complex imputation procedures make sense except with substantially higher number of studies in than is common in meta-analyses in psychology.

These are practical concerns with applying reliability corrections, but a more pernicious challenge to treating disattenuated effect sizes as equivalent to construct scores is the problem of validity (Borsboom & Mellenbergh, 2002; Hunter & Schmidt, 2015, p. 41 – 53, 2nd edition). First, doing so requires assuming that the constructs underlying measurements are unidimensional in nature (Borsboom & Mellenbergh, 2002), which in many cases is a strong assumption. Second, to treat disattenuated effect sizes as latent construct scores requires assuming that the applied measurement instruments measure the construct of interest directly rather than through some proxy variable. Although corrections akin to the classic unreliability correction can be applied for simple causal models, these methods quickly become insufficient as the relationship between latent constructs grows in complexity (Hunter & Schmidt, 2015, p. 41 – 53, 2nd edition). As such, some argue that simple corrections for unreliability under the classical test theory framework are untenable and potentially misleading if one’s interest lies in latent constructs (Borsboom & Mellenbergh, 2002). From this perspective, researchers should consider modern psychometric models such as Structural Equation Models (SEM; e.g., REF) which permit delving deeper into measurements and assumptions about latent construct relationships.

Meta-analysis can be conceptualized as a SEM model, which holds promise that the benefits of modern psychometric theory might be applied also when the interest is in summarizing research across multiple studies. In Meta-Analytic Structural Equation Modeling (MASEM; e.g., Cheung & Chan, 2005) unreliability in primary studies can be corrected in three ways (Gnambs & Sengewald, 2023): 1) by estimating a SEM model in each primary study and then pooling estimates (parameter-based MASEM: Cheung, 2015, p. 241), 2) by directly modeling latent variables in the MASEM, or 3) by first correcting correlations before introducing them into the MASEM framework (i.e., fundamentally identical to the classical test theory approach). The first option is limited because it is rarely the case that the same SEM model can be fit across multiple independent datasets, except in highly structure cases (e.g., Brunner et al., 2022). The second option requires item-level statistics rather than summary values, which are rarely available to meta-analysts. The third option, naturally, suffers the same challenges and assumptions as corrections under classical test theory. As such, the potential to use MASEM to take advantages of modern psychometric theory when correcting for unreliability is challenging in practice. That said, multi-lab replication projects, which are increasingly common in psychology (e.g., REF1, REF2), are well-suited to take advantage of MASEM.

In the end, meta-analysis depends on the quality of the primary studies that make up its data. As long as researchers in psychology underappreciate the importance of measurement, meta-analytic estimates will be poor. For meta-analysts that consider acceptable the challenges and assumptions of corrections to unreliability, the many caveats inherent in doing so should be acknowledged and estimates based on both corrected and uncorrected estimates reported. Meta-analysts who consider corrections untenable need to consider that heterogeneity estimates may be severely underestimated. In either case, meta-analysts should get used to extracting reliability information whenever available from primary studies. The good news is that large true heterogeneity ($\tau > 0.1$) is on average distinguishable from zero or small estimates even in observed heterogeneity estimates.

## Limitations

In our simulations we assumed equal reliability for the measures $X$ and $Y$ (i.e., $R_i = R_{xx'} = R_{yy'}$) within a study, which is not realistic. However, as we never use the reliabilities individually (i.e., only expressed as $\sqrt{R_{xx'}} \times \sqrt{R_{yy'}}$) we do not expect major impact on our results. Consider that if $R_i = 0.7$ then $\sqrt{R_{xx'}} \times \sqrt{R_{yy'}} = 0.7$ whereas if if $R_{xx'}$ and $R_{yy'}$ took other values from a distribution centered on 0.7 this value would be lower (e.g., if $R_{xx'} = 0.5$ and $R_{yy'} = 0.9$ then $\sqrt{R_{xx'}} \times \sqrt{R_{yy'}} \approx 0.67$). As we use the results of $\sqrt{R_{xx'}} \times \sqrt{R_{yy'}}$ in our simulations to generating observed effect sizes from true effect sizes (i.e., $r_{xy} = \rho_{xy} \times \sqrt{R_{xx'}} \times \sqrt{R_{yy'}}$, where $\rho_{xy}$ is the true effect size after sampling error), letting the reliability for $X$ and $Y$ vary would lead to slightly larger attenuation of observed effect sizes on average. That is, letting the reliability of $X$ and $Y$ vary within studies would lead to larger negative bias in observed heterogeneity estimates than in our simulations, indicating that the effect might even be somewhat more severe than in our estimates.

We also only considered bias in heterogeneity due to unreliability in primary studies, but there exist other measurement artifacts such as restriction of range, dichotomization and more (Hunter & Schmidt, 2015). Although measurement error is the only measurement artifact that is always present, many of these other measurement artifacts may impact heterogeneity estimates when they are present. Based on Hunter & Schmidt’s arguments that measurement artifacts all tend to attenuate effect sizes in a similar way to unreliability, we expect that their net effect on heterogeneity estimates may similarly be a negative bias. The extent to which this prediction is true (and whether these measurement artifacts can be corrected for in MASEM; Gnambs & Sengewalds, 2023) awaits further research.

We also used a fixed sample size (N) across studies within a meta-analysis in our simulations, which is unrealistic. We showed in Supplement XX that increasing the fixed sample size led to slightly better observed heterogeneity estimates for low reliability. Implementing a distribution of sample sizes from which within meta-analyses sample sizes was drawn (e.g., as done by REF1; ref2) could be one way to increase realism in sample sizes, although to the detriment of interpretability. However, given a sufficiently large number of repetitions in our simulations, this would not change estimates of bias, assuming the same mean sample sizes were used. That said, moving from fixed sample sizes to a distribution of sample sizes would affect the variability of any estimates, but this was not the focus of our study.

We did not examine variance in heterogeneity estimates, or their root mean squared error. Our focus was on to what extent unreliability leads to bias in heterogeneity estimates and should be a concern for substantive researcher attempting to use meta-analysis to learn more about the contextual sensitivity of their research topic. As such, although we apply different meta-analytic models and effect sizes, the purpose was not to compare the efficiency of these different modes of estimation, but to show our results with the dominant meta-analytic approaches for correlations and demonstrate that they are not to any large degree dependent on the meta-analytic approach. As the effect of unreliability in primary studies is simply to change the heterogeneity of effect sizes the results of previous studies on the comparable performance of different heterogeneity estimators applies also here (e.g., REF1, REF2, REF3). That said, examining the variance of heterogeneity estimates could be useful to gain a better idea of how likely a particular observed heterogeneity estimate is to belong to different distributions of true heterogeneity. As this would depend on N, K, mean reliability, variability in reliability, average effect size, observed heterogeneity, and the true heterogeneity levels being compared, implementing such an analysis would probably be best done on a case-by-case basis (e.g., through a webapp such as developed for sensitivity analyses of the effect of publication bias on heterogeneity by Augusteijn et al., YEAR). Given the challenges in correcting for unreliability, such a sensitivity analysis could be a promising tool to reason around the impact of unreliability in applied meta-analysis.

Our simulations assume reliabilities to be known. This is never the case, and any estimate of reliability is at best a lower bound estimate. By assuming reliabilities to be known, we were able to show the bias of observed heterogeneity estimates compared to the ideal case of no measurement error. However, this may mislead readers to believe that a simple correction for unreliability will leave them with an unbiased estimate. As discussed in the section ‘should meta-analysts correct for unreliability?’ correcting for reliability based on estimated reliability is not straightforward and whether the resulting estimate will be unbiased is contingent on many factors and in many cases unresolved. As such, examining the effect of corrections based on estimated reliabilities was beyond the scope of this study.

## Conclusion

Imperfect measurement reliability in primary studies generally leads to a severe underestimate of observed meta-analytic heterogeneity. As few meta-analyses in psychology correct for unreliability in primary studies, heterogeneity is likely underestimated in a large proportion of meta-analyses in psychology. Consequently, variability in effects has been ignored, moderators potentially neglected and theoretical development in psychology been misguided by meta-analyses. Sophisticated methods (MASEM) for correcting measurement unreliability are only applicable in exceptional cases and corrections based on classical test theory come with caveats and assumptions. Accurate estimation of meta-analytic heterogeneity is difficult and will remain so unless measurement concerns (REF) are taken seriously in primary research. The good news is that study designs (multi-lab replication studies) which can apply more sophisticated versions of MASEM are becoming increasingly common in psychology, and that large true heterogeneity can on average be distinguished from zero or small heterogeneity even in the absence of corrections to unreliability in primary studies.
